---
title: "zarrs Backend"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{zarrs Backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
sample_dir <- tools::R_user_dir("pizzarr")
clean <- !dir.exists(sample_dir)
```

pizzarr ships in two tiers. The CRAN build is pure R --- no Rust compilation,
no system dependencies. It handles local and HTTP Zarr stores with sequential
chunk I/O via `lapply`. The r-universe build compiles in the
[zarrs](https://github.com/zarrs/zarrs) Rust crate via
[extendr](https://extendr.github.io/rextendr/), adding parallel decompression,
cloud-native store backends (S3, GCS), and codecs beyond what R packages provide.

The split exists because CRAN's macOS build machines ship a Rust toolchain
(rustc 1.84) that is too old for zarrs, which requires rustc >= 1.91.
r-universe builds against the latest stable toolchain, so it can compile zarrs
and distribute pre-built binaries. End users on either tier install with
`install.packages()` --- no Rust toolchain needed.

## Checking availability

```{r}
library(pizzarr)

has_zarrs <- pizzarr:::.pizzarr_env$zarrs_available
```

`pizzarr_compiled_features()` lists the feature flags compiled into the zarrs
backend. On the CRAN tier it returns `character(0)` with a message; on the
r-universe tier it returns the compiled capabilities:

```{r}
pizzarr_compiled_features()
```

The internal flag `.pizzarr_env$zarrs_available` is a logical scalar set once
at package load. Dispatch logic throughout pizzarr checks this flag to decide
whether to call into Rust or fall through to the R-native path:

```{r}
pizzarr:::.pizzarr_env$zarrs_available
```

## Upgrading to the zarrs tier

`pizzarr_upgrade()` prints the r-universe install command when zarrs is not
compiled in, or confirms that the backend is already present:

```{r}
pizzarr_upgrade()
```

The startup message that CRAN users see on `library(pizzarr)` can be silenced
with `options(pizzarr.suggest_runiverse = FALSE)`.

## Probing store metadata

The examples below require the zarrs backend. When this vignette is built
without it, the code chunks are not evaluated.

`zarrs_node_exists()` opens a filesystem store via the Rust backend, probes
for V2 and V3 metadata keys at a given path, and returns a list with three
fields: `exists` (logical), `node_type` (character), and `zarr_format`
(integer or NULL). The store handle is cached on the Rust side --- subsequent
calls to the same store path reuse it without re-opening.

### V2 store

```{r, eval=has_zarrs}
v2_root <- pizzarr_sample("fixtures/v2/data.zarr")

# Root group
zarrs_node_exists(v2_root, "")
```

```{r, eval=has_zarrs}
# An array within the store
zarrs_node_exists(v2_root, "1d.contiguous.lz4.i2")
```

```{r, eval=has_zarrs}
# A path that does not exist
zarrs_node_exists(v2_root, "does_not_exist")
```

### V3 store

V2 and V3 detection is automatic. zarrs probes for `zarr.json` first (V3),
then falls back to `.zarray` / `.zgroup` (V2):

```{r, eval=has_zarrs}
v3_root <- pizzarr_sample("fixtures/v3/data.zarr")

zarrs_node_exists(v3_root, "")
```

## Store cache management

The Rust backend holds open store handles in a process-global cache keyed by
normalized path. `zarrs_close_store()` removes a handle from the cache and
returns `TRUE`. A second call to the same path returns `FALSE` --- it was
already removed:

```{r, eval=has_zarrs}
zarrs_close_store(v2_root)
zarrs_close_store(v2_root)
```

```{r, eval=has_zarrs}
zarrs_close_store(v3_root)
```

## Reading data via zarrs

<!-- TODO: Phase 2 — zarrs_retrieve_subset(), parallel chunk reads -->

## Writing data via zarrs

<!-- TODO: Phase 3 — zarrs_store_subset(), zarrs_create_array() -->

## Remote stores

<!-- TODO: Phase 4 — HTTP, S3, GCS backends via object_store -->

```{r, include=FALSE}
if (clean) unlink(sample_dir, recursive = TRUE)
```
