---
title: "zarrs Backend"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{zarrs Backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
sample_dir <- tools::R_user_dir("pizzarr")
clean <- !dir.exists(sample_dir)
```

pizzarr ships in two tiers. The CRAN build is pure R --- no Rust compilation,
no system dependencies. It handles local and HTTP Zarr stores with sequential
chunk I/O via `lapply`. The r-universe build compiles in the
[zarrs](https://github.com/zarrs/zarrs) Rust crate via
[extendr](https://extendr.github.io/rextendr/), adding parallel decompression,
cloud-native store backends (S3, GCS), and codecs beyond what R packages provide.

The split exists because CRAN's macOS build machines ship a Rust toolchain
(rustc 1.84) that is too old for zarrs, which requires rustc >= 1.91.
r-universe builds against the latest stable toolchain, so it can compile zarrs
and distribute pre-built binaries. End users on either tier install with
`install.packages()` --- no Rust toolchain needed.

## Checking availability

```{r}
library(pizzarr)

has_zarrs <- pizzarr:::.pizzarr_env$zarrs_available
```

`pizzarr_compiled_features()` lists the feature flags compiled into the zarrs
backend. On the CRAN tier it returns `character(0)` with a message; on the
r-universe tier it returns the compiled capabilities:

```{r}
pizzarr_compiled_features()
```

The internal flag `.pizzarr_env$zarrs_available` is a logical scalar set once
at package load. Dispatch logic throughout pizzarr checks this flag to decide
whether to call into Rust or fall through to the R-native path:

```{r}
pizzarr:::.pizzarr_env$zarrs_available
```

## Upgrading to the zarrs tier

`pizzarr_upgrade()` prints the r-universe install command when zarrs is not
compiled in, or confirms that the backend is already present:

```{r}
pizzarr_upgrade()
```

The startup message that CRAN users see on `library(pizzarr)` can be silenced
with `options(pizzarr.suggest_runiverse = FALSE)`.

## Probing store metadata

The examples below require the zarrs backend. When this vignette is built
without it, the code chunks are not evaluated.

`zarrs_node_exists()` opens a filesystem store via the Rust backend, probes
for V2 and V3 metadata keys at a given path, and returns a list with three
fields: `exists` (logical), `node_type` (character), and `zarr_format`
(integer or NULL). The store handle is cached on the Rust side --- subsequent
calls to the same store path reuse it without re-opening.

### V2 store

```{r, eval=has_zarrs}
v2_root <- pizzarr_sample("fixtures/v2/data.zarr")

# Root group
zarrs_node_exists(v2_root, "")
```

```{r, eval=has_zarrs}
# An array within the store
zarrs_node_exists(v2_root, "1d.contiguous.lz4.i2")
```

```{r, eval=has_zarrs}
# A path that does not exist
zarrs_node_exists(v2_root, "does_not_exist")
```

### V3 store

V2 and V3 detection is automatic. zarrs probes for `zarr.json` first (V3),
then falls back to `.zarray` / `.zgroup` (V2):

```{r, eval=has_zarrs}
v3_root <- pizzarr_sample("fixtures/v3/data.zarr")

zarrs_node_exists(v3_root, "")
```

## Store cache management

The Rust backend holds open store handles in a process-global cache keyed by
normalized path. `zarrs_close_store()` removes a handle from the cache and
returns `TRUE`. A second call to the same path returns `FALSE` --- it was
already removed:

```{r, eval=has_zarrs}
zarrs_close_store(v2_root)
zarrs_close_store(v2_root)
```

```{r, eval=has_zarrs}
zarrs_close_store(v3_root)
```

## Array metadata

`zarrs_open_array_metadata()` opens a zarrs array and returns its metadata as a
named list. The store handle is cached, so repeated calls to the same store are
fast. The returned list contains `shape`, `chunks`, `dtype`, `r_type`,
`fill_value_json`, `zarr_format`, and `order`.

### V2 array

```{r, eval=has_zarrs}
v2_root <- pizzarr_sample("fixtures/v2/data.zarr")
zarrs_open_array_metadata(v2_root, "1d.contiguous.raw.i2")
```

### V3 array

V3 arrays work the same way. The `zarr_format` field distinguishes V2 from V3:

```{r, eval=has_zarrs}
v3_root <- pizzarr_sample("fixtures/v3/data.zarr")
zarrs_open_array_metadata(v3_root, "1d.contiguous.gzip.i2")
```

### Data type classification

The `r_type` field maps zarrs data types to R-compatible type families. zarrs
numeric types are classified as `"double"`, `"integer"`, or `"logical"` based on
what R can represent natively:

- **double**: float64 (zero-cost), float32 (widened), uint32/int64/uint64 (widened, precision risk > 2^53)
- **integer**: int32 (zero-cost), int8/int16/uint8/uint16 (widened)
- **logical**: bool

Unsupported types (strings, complex) report `"unsupported"` and fall back to the
R-native code path.

```{r, eval=has_zarrs}
zarrs_close_store(v2_root)
zarrs_close_store(v3_root)
```

## Runtime info and tuning

`zarrs_runtime_info()` reports the current zarrs configuration --- the codec
concurrency target, how many store handles are cached, and which features were
compiled in:

```{r, eval=has_zarrs}
zarrs_runtime_info()
```

`zarrs_set_codec_concurrent_target()` controls how many concurrent codec
operations zarrs uses within a single array operation. The default matches the
number of available CPUs. Setting it to 1 disables parallel decompression:

```{r, eval=has_zarrs}
zarrs_set_codec_concurrent_target(2L)
zarrs_runtime_info()$codec_concurrent_target
```

## Reading data via zarrs

When the zarrs backend is available and the selection is a contiguous slice
(step == 1), `ZarrArray$get_item()` dispatches reads to zarrs automatically.
zarrs handles chunk identification, parallel decompression, and codec execution
internally, bypassing pizzarr's R-native chunk loop. Scalar integer selections
(e.g., selecting a single row of a matrix) are also eligible --- they become
length-1 ranges on the Rust side. Unsupported selections (step > 1 slices,
fancy indexing, MemoryStore) fall through to the R-native path transparently.

### Basic read

```{r, eval=has_zarrs}
d <- tempfile("zarrs_vignette_")
z <- zarr_create(store = d, shape = c(100L, 50L), chunks = c(10L, 10L),
                 dtype = "<f8")
z$set_item("...", array(as.double(seq_len(5000)), dim = c(100, 50)))

# Re-open and read a subset --- zarrs handles the chunk I/O
z2 <- zarr_open(store = d)
result <- z2$get_item(list(slice(1L, 10L), slice(1L, 5L)))
dim(result$data)
```

### Direct zarrs_get_subset call

For lower-level access, `zarrs_get_subset()` reads a contiguous subset
directly via the Rust backend. Ranges are 0-based with exclusive stop, matching
zarrs conventions:

```{r, eval=has_zarrs}
result <- zarrs_get_subset(d, "", list(c(0L, 10L), c(0L, 5L)), NULL)
str(result)
```

### Concurrency control

The optional `concurrent_target` parameter (or the `pizzarr.concurrent_target`
R option) controls how many parallel codec operations zarrs uses within a single
read call. Setting it to `1L` disables parallel decompression:

```{r, eval=has_zarrs}
result <- zarrs_get_subset(d, "", list(c(0L, 10L), c(0L, 5L)), 1L)
length(result$data)
```

```{r, eval=has_zarrs}
zarrs_close_store(d)
unlink(d, recursive = TRUE)
```

## Writing data via zarrs

The write path mirrors the read path. When the zarrs backend is available and
the selection qualifies (contiguous slices, filesystem-backed store),
`ZarrArray$set_item()` dispatches writes to zarrs instead of iterating over
chunks in R. zarrs encodes the data, splits it across the affected chunks, and
writes them to disk --- using its internal thread pool for parallel compression
when multiple chunks are involved.

Data type narrowing happens on the Rust side. R doubles narrow to the array's
stored type (float32, int64, uint32, etc.) and R integers narrow to smaller
integer types (int16, int8, uint8, uint16) with range checking. An out-of-range
value produces an error rather than silent truncation.

### Basic write

```{r, eval=has_zarrs}
d <- tempfile("zarrs_write_vignette_")
z <- zarr_create(store = d, shape = c(20L, 10L), chunks = c(10L, 10L),
                 dtype = "<f8")

# set_item dispatches to zarrs when eligible
z$set_item("...", array(as.double(1:200), dim = c(20, 10)))

# Read back to confirm
z2 <- zarr_open(store = d)
result <- z2$get_item(list(slice(1L, 5L), slice(1L, 3L)))
result$data
```

### Partial overwrite

Writing to a subset of an existing array works the same way. zarrs reads the
affected chunks, merges the new data, and writes them back:

```{r, eval=has_zarrs}
# Overwrite rows 3-7, columns 1-2
z$set_item(list(slice(3L, 7L), slice(1L, 2L)),
           array(rep(-1.0, 10), dim = c(5, 2)))

result <- z2$get_item(list(slice(1L, 10L), slice(1L, 3L)))
result$data
```

### Direct zarrs_set_subset call

`zarrs_set_subset()` provides lower-level access to the Rust write path.
Data is a flat vector in R's native F-order (column-major) --- the Rust
backend handles the F-to-C order conversion internally. The function returns
`TRUE` on success:

```{r, eval=has_zarrs}
# Write 10 values to the first row (0-based range [0, 1) x [0, 10))
zarrs_set_subset(d, "", list(c(0L, 1L), c(0L, 10L)),
                   as.double(101:110), NULL)

result <- zarrs_get_subset(d, "", list(c(0L, 1L), c(0L, 10L)), NULL)
result$data
```

```{r, eval=has_zarrs}
zarrs_close_store(d)
unlink(d, recursive = TRUE)
```

## HTTP reads via zarrs

When the `http_sync` feature is compiled in, the zarrs backend can read
directly from HTTP/HTTPS Zarr stores using the
[zarrs_http](https://github.com/zarrs/zarrs_http) crate. This bypasses
pizzarr's R-native `crul`-based chunk loop, giving parallel chunk decode
on remote data.

HTTP stores are read-only in zarrs --- write dispatch (`set_item`) falls
through to the R-native path automatically.

### Transparent dispatch

The zarrs fast path activates automatically when an `HttpStore`-backed
array is read with a contiguous selection. No code changes are needed
compared to the R-native path:

```{r, eval=has_zarrs && ("http_sync" %in% pizzarr_compiled_features())}
url <- "https://raw.githubusercontent.com/DOI-USGS/rnz/main/inst/extdata/bcsd.zarr"

z <- zarr_open(store = HttpStore$new(url))

# zarrs handles the HTTP reads + parallel decompression
pr <- z$get_item("pr")
pr
```

```{r, eval=has_zarrs && ("http_sync" %in% pizzarr_compiled_features())}
# Read a subset --- zarrs fetches only the chunks that overlap
result <- pr$get_item(list(slice(1L, 3L), slice(1L, 5L), slice(1L, 5L)))
dim(result$data)
```

### Direct zarrs_get_subset from HTTP

`zarrs_get_subset()` also works with HTTP URLs. The store handle is cached
on the Rust side, so repeated reads to the same URL reuse the connection:

```{r, eval=has_zarrs && ("http_sync" %in% pizzarr_compiled_features())}
meta <- zarrs_open_array_metadata(url, "pr")
str(meta[c("shape", "dtype", "zarr_format")])
```

```{r, eval=has_zarrs && ("http_sync" %in% pizzarr_compiled_features())}
# Read a single element (first along each dimension)
ranges <- lapply(seq_along(meta$shape), function(i) c(0L, 1L))
result <- zarrs_get_subset(url, "pr", ranges, NULL)
result$data
```

```{r, eval=has_zarrs && ("http_sync" %in% pizzarr_compiled_features())}
zarrs_close_store(url)
```

### Feature detection

Check whether HTTP support is compiled in with `pizzarr_compiled_features()`.
When `"http_sync"` is present, zarrs can open `http://` and `https://` URLs.
When it is absent, HTTP reads fall through to the R-native `crul`-based path:

```{r, eval=has_zarrs}
"http_sync" %in% pizzarr_compiled_features()
```

### Future: S3 and GCS

Cloud object-store backends (S3, GCS, Azure) via
[zarrs_object_store](https://github.com/zarrs/zarrs_object_store) are planned
for a future iteration. These require a tokio async runtime and the `full`
Cargo feature flag.

## C/F order handling

zarrs stores data in C-order (row-major), while R uses F-order (column-major).
The Rust backend handles this conversion transparently:

- **Reads:** `zarrs_get_subset()` returns data in F-order, ready for
  `array(data, dim = shape)` with no `aperm()` needed.
- **Writes:** `zarrs_set_subset()` accepts F-order data and converts to C-order
  internally before writing to the store.

The transpose uses cache-blocked tiling for 2D arrays and output-order
iteration with incremental index tracking for higher dimensions, matching or
exceeding the performance of R's C-level `aperm()`.

```{r, include=FALSE}
if (clean) unlink(sample_dir, recursive = TRUE)
```
