---
title: "zarrs Backend"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{zarrs Backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
sample_dir <- tools::R_user_dir("pizzarr")
clean <- !dir.exists(sample_dir)
```

pizzarr ships in two tiers. The CRAN build is pure R --- no Rust compilation,
no system dependencies. It handles local and HTTP Zarr stores with sequential
chunk I/O via `lapply`. The r-universe build compiles in the
[zarrs](https://github.com/zarrs/zarrs) Rust crate via
[extendr](https://extendr.github.io/rextendr/), adding parallel decompression,
cloud-native store backends (S3, GCS), and codecs beyond what R packages provide.

The split exists because CRAN's macOS build machines ship a Rust toolchain
(rustc 1.84) that is too old for zarrs, which requires rustc >= 1.91.
r-universe builds against the latest stable toolchain, so it can compile zarrs
and distribute pre-built binaries. End users on either tier install with
`install.packages()` --- no Rust toolchain needed.

## Checking availability

```{r}
library(pizzarr)

has_zarrs <- pizzarr:::.pizzarr_env$zarrs_available
```

`pizzarr_compiled_features()` lists the feature flags compiled into the zarrs
backend. On the CRAN tier it returns `character(0)` with a message; on the
r-universe tier it returns the compiled capabilities:

```{r}
pizzarr_compiled_features()
```

The internal flag `.pizzarr_env$zarrs_available` is a logical scalar set once
at package load. Dispatch logic throughout pizzarr checks this flag to decide
whether to call into Rust or fall through to the R-native path:

```{r}
pizzarr:::.pizzarr_env$zarrs_available
```

## Upgrading to the zarrs tier

`pizzarr_upgrade()` prints the r-universe install command when zarrs is not
compiled in, or confirms that the backend is already present:

```{r}
pizzarr_upgrade()
```

The startup message that CRAN users see on `library(pizzarr)` can be silenced
with `options(pizzarr.suggest_runiverse = FALSE)`.

## Probing store metadata

The examples below require the zarrs backend. When this vignette is built
without it, the code chunks are not evaluated.

`zarrs_node_exists()` opens a filesystem store via the Rust backend, probes
for V2 and V3 metadata keys at a given path, and returns a list with three
fields: `exists` (logical), `node_type` (character), and `zarr_format`
(integer or NULL). The store handle is cached on the Rust side --- subsequent
calls to the same store path reuse it without re-opening.

### V2 store

```{r, eval=has_zarrs}
v2_root <- pizzarr_sample("fixtures/v2/data.zarr")

# Root group
zarrs_node_exists(v2_root, "")
```

```{r, eval=has_zarrs}
# An array within the store
zarrs_node_exists(v2_root, "1d.contiguous.lz4.i2")
```

```{r, eval=has_zarrs}
# A path that does not exist
zarrs_node_exists(v2_root, "does_not_exist")
```

### V3 store

V2 and V3 detection is automatic. zarrs probes for `zarr.json` first (V3),
then falls back to `.zarray` / `.zgroup` (V2):

```{r, eval=has_zarrs}
v3_root <- pizzarr_sample("fixtures/v3/data.zarr")

zarrs_node_exists(v3_root, "")
```

## Store cache management

The Rust backend holds open store handles in a process-global cache keyed by
normalized path. `zarrs_close_store()` removes a handle from the cache and
returns `TRUE`. A second call to the same path returns `FALSE` --- it was
already removed:

```{r, eval=has_zarrs}
zarrs_close_store(v2_root)
zarrs_close_store(v2_root)
```

```{r, eval=has_zarrs}
zarrs_close_store(v3_root)
```

## Array metadata

`zarrs_open_array_metadata()` opens a zarrs array and returns its metadata as a
named list. The store handle is cached, so repeated calls to the same store are
fast. The returned list contains `shape`, `chunks`, `dtype`, `r_type`,
`fill_value_json`, `zarr_format`, and `order`.

### V2 array

```{r, eval=has_zarrs}
v2_root <- pizzarr_sample("fixtures/v2/data.zarr")
zarrs_open_array_metadata(v2_root, "1d.contiguous.raw.i2")
```

### V3 array

V3 arrays work the same way. The `zarr_format` field distinguishes V2 from V3:

```{r, eval=has_zarrs}
v3_root <- pizzarr_sample("fixtures/v3/data.zarr")
zarrs_open_array_metadata(v3_root, "1d.contiguous.gzip.i2")
```

### Data type classification

The `r_type` field maps zarrs data types to R-compatible type families. zarrs
numeric types are classified as `"double"`, `"integer"`, or `"logical"` based on
what R can represent natively:

- **double**: float64 (zero-cost), float32 (widened), uint32/int64/uint64 (widened, precision risk > 2^53)
- **integer**: int32 (zero-cost), int8/int16/uint8/uint16 (widened)
- **logical**: bool

Unsupported types (strings, complex) report `"unsupported"` and fall back to the
R-native code path.

```{r, eval=has_zarrs}
zarrs_close_store(v2_root)
zarrs_close_store(v3_root)
```

## Runtime info and tuning

`zarrs_runtime_info()` reports the current zarrs configuration --- the codec
concurrency target, thread pool size, how many store handles are cached, and
which features were compiled in:

```{r, eval=has_zarrs}
zarrs_runtime_info()
```

### pizzarr_config()

`pizzarr_config()` is the main interface for viewing and changing concurrency
settings. Called with no arguments it returns the current state; with arguments
it sets the specified values:

```{r, eval=has_zarrs}
# View current settings
pizzarr_config()

# Set codec concurrency to 2 parallel operations per read/write
pizzarr_config(concurrent_target = 2L)
zarrs_runtime_info()$codec_concurrent_target
```

Three settings are available:

- **nthreads** --- rayon thread pool size. Set-once per R session (the thread
  pool can only be initialised once). For reliable session-level control, set
  the `PIZZARR_NTHREADS` environment variable before starting R.
- **concurrent_target** --- how many codec operations zarrs runs in parallel
  within a single read or write call. Can be changed at any time.
- **http_batch_range_requests** --- whether HTTP stores use multipart range
  requests (default TRUE). Set to FALSE for servers with incomplete multipart
  support. Takes effect on the next store open.

All three settings can also be configured via environment variables
(`PIZZARR_NTHREADS`, `PIZZARR_CONCURRENT_TARGET`,
`PIZZARR_HTTP_BATCH_RANGE_REQUESTS`) or R options (`pizzarr.nthreads`, etc.),
which are read at package load time. Environment variables persist across
sessions without needing `.Rprofile` edits.

The lower-level `zarrs_set_codec_concurrent_target()` function is still
available for direct use:

```{r, eval=has_zarrs}
zarrs_set_codec_concurrent_target(2L)
zarrs_runtime_info()$codec_concurrent_target
```

## Reading data via zarrs

When the zarrs backend is available and the selection is a contiguous slice
(step == 1), `ZarrArray$get_item()` dispatches reads to zarrs automatically.
zarrs handles chunk identification, parallel decompression, and codec execution
internally, bypassing pizzarr's R-native chunk loop. Scalar integer selections
(e.g., selecting a single row of a matrix) are also eligible --- they become
length-1 ranges on the Rust side. Unsupported selections (step > 1 slices,
fancy indexing, MemoryStore) fall through to the R-native path transparently.

### Basic read

```{r, eval=has_zarrs}
d <- tempfile("zarrs_vignette_")
z <- zarr_create(store = d, shape = c(100L, 50L), chunks = c(10L, 10L),
                 dtype = "<f8")
z$set_item("...", array(as.double(seq_len(5000)), dim = c(100, 50)))

# Re-open and read a subset --- zarrs handles the chunk I/O
z2 <- zarr_open(store = d)
result <- z2$get_item(list(slice(1L, 10L), slice(1L, 5L)))
dim(result$data)
```

### Direct zarrs_get_subset call

For lower-level access, `zarrs_get_subset()` reads a contiguous subset
directly via the Rust backend. Ranges are 0-based with exclusive stop, matching
zarrs conventions:

```{r, eval=has_zarrs}
result <- zarrs_get_subset(d, "", list(c(0L, 10L), c(0L, 5L)), NULL)
str(result)
```

### Concurrency control

The optional `concurrent_target` parameter (or the `pizzarr.concurrent_target`
R option) controls how many parallel codec operations zarrs uses within a single
read call. Setting it to `1L` disables parallel decompression:

```{r, eval=has_zarrs}
result <- zarrs_get_subset(d, "", list(c(0L, 10L), c(0L, 5L)), 1L)
length(result$data)
```

```{r, eval=has_zarrs}
zarrs_close_store(d)
unlink(d, recursive = TRUE)
```

## Creating arrays via zarrs

When the zarrs backend is available and the store is a writable filesystem path,
`zarr_create()` dispatches array creation to zarrs instead of building metadata
JSON in R. zarrs validates the metadata structure, writes it to the store, and
the array is ready for data. The dispatch is transparent --- the same
`zarr_create()` call works on both tiers, and unsupported configurations
(MemoryStore, object dtypes, custom filters) fall through to the R-native path.

### Transparent dispatch

The `zarr_create()` examples earlier in this vignette already use this path when
zarrs is available. The zarrs backend handles V2 and V3 formats, all 11 numeric
data types, and four codec presets:

```{r, eval=has_zarrs}
# V3 array with gzip compression
d <- tempfile("zarrs_create_vignette_")
z <- zarr_create(store = d, shape = c(20L, 10L), chunks = c(10L, 10L),
                 dtype = "<f8", zarr_format = 3L)
z
```

```{r, eval=has_zarrs}
# Confirm V3 metadata was written
file.exists(file.path(d, "zarr.json"))
```

```{r, eval=has_zarrs}
zarrs_close_store(d)
unlink(d, recursive = TRUE)
```

### Direct zarrs_create_array call

`zarrs_create_array()` provides lower-level access to the Rust creation path.
It accepts V3-style data type names (`"float64"`, `"int32"`, `"bool"`, etc.)
and a codec preset string (`"none"`, `"gzip"`, `"blosc"`, or `"zstd"`).
The return value is the same metadata list as `zarrs_open_array_metadata()`:

```{r, eval=has_zarrs}
d <- tempfile("zarrs_create_direct_")
dir.create(d)

meta <- zarrs_create_array(
  store_url = d,
  array_path = "",
  shape = c(100L, 50L),
  chunks = c(10L, 10L),
  dtype = "float64",
  codec_preset = "gzip",
  fill_value = 0.0,
  attributes_json = "{}",
  zarr_format = 3L
)
str(meta)
```

The array is immediately usable for reads and writes:

```{r, eval=has_zarrs}
zarrs_set_subset(d, "", list(c(0L, 10L), c(0L, 5L)),
                 as.double(1:50), NULL)
result <- zarrs_get_subset(d, "", list(c(0L, 10L), c(0L, 5L)), NULL)
head(result$data)
```

```{r, eval=has_zarrs}
zarrs_close_store(d)
unlink(d, recursive = TRUE)
```

### Codec presets

The zarrs creation path supports four named codec presets. Custom codec
configurations fall through to the R-native path.

| Preset | V2 compressor | V3 codec chain | Notes |
|--------|--------------|----------------|-------|
| `"none"` | null | bytes only | No compression |
| `"gzip"` | gzip, level 1 | bytes + gzip(1) | Fast, reasonable ratio |
| `"blosc"` | blosc, lz4, clevel 5 | bytes + blosc(lz4, 5) | Requires `blosc` feature |
| `"zstd"` | --- | bytes + zstd(3) | V3 only; requires `zstd` feature |

One difference from the R-native path: zarrs uses the `"gzip"` compressor id
for V2 arrays, while zarr-python uses `"zlib"`. Both produce gzip-compatible
output, and zarrs reads either id when opening existing arrays.

## Writing data via zarrs

The write path mirrors the read path. When the zarrs backend is available and
the selection qualifies (contiguous slices, filesystem-backed store),
`ZarrArray$set_item()` dispatches writes to zarrs instead of iterating over
chunks in R. zarrs encodes the data, splits it across the affected chunks, and
writes them to disk --- using its internal thread pool for parallel compression
when multiple chunks are involved.

Data type narrowing happens on the Rust side. R doubles narrow to the array's
stored type (float32, int64, uint32, etc.) and R integers narrow to smaller
integer types (int16, int8, uint8, uint16) with range checking. An out-of-range
value produces an error rather than silent truncation.

### Basic write

```{r, eval=has_zarrs}
d <- tempfile("zarrs_write_vignette_")
z <- zarr_create(store = d, shape = c(20L, 10L), chunks = c(10L, 10L),
                 dtype = "<f8")

# set_item dispatches to zarrs when eligible
z$set_item("...", array(as.double(1:200), dim = c(20, 10)))

# Read back to confirm
z2 <- zarr_open(store = d)
result <- z2$get_item(list(slice(1L, 5L), slice(1L, 3L)))
result$data
```

### Partial overwrite

Writing to a subset of an existing array works the same way. zarrs reads the
affected chunks, merges the new data, and writes them back:

```{r, eval=has_zarrs}
# Overwrite rows 3-7, columns 1-2
z$set_item(list(slice(3L, 7L), slice(1L, 2L)),
           array(rep(-1.0, 10), dim = c(5, 2)))

result <- z2$get_item(list(slice(1L, 10L), slice(1L, 3L)))
result$data
```

### Direct zarrs_set_subset call

`zarrs_set_subset()` provides lower-level access to the Rust write path.
Data is a flat vector in R's native F-order (column-major) --- the Rust
backend handles the F-to-C order conversion internally. The function returns
`TRUE` on success:

```{r, eval=has_zarrs}
# Write 10 values to the first row (0-based range [0, 1) x [0, 10))
zarrs_set_subset(d, "", list(c(0L, 1L), c(0L, 10L)),
                   as.double(101:110), NULL)

result <- zarrs_get_subset(d, "", list(c(0L, 1L), c(0L, 10L)), NULL)
result$data
```

```{r, eval=has_zarrs}
zarrs_close_store(d)
unlink(d, recursive = TRUE)
```

## HTTP reads via zarrs

When the `http_sync` feature is compiled in, the zarrs backend can read
directly from HTTP/HTTPS Zarr stores using the
[zarrs_http](https://github.com/zarrs/zarrs_http) crate. This bypasses
pizzarr's R-native `crul`-based chunk loop, giving parallel chunk decode
on remote data.

HTTP stores are read-only in zarrs --- write dispatch (`set_item`) falls
through to the R-native path automatically.

### Transparent dispatch

The zarrs fast path activates automatically when an `HttpStore`-backed
array is read with a contiguous selection. No code changes are needed
compared to the R-native path:

```{r, eval=has_zarrs && ("http_sync" %in% pizzarr_compiled_features())}
url <- "https://raw.githubusercontent.com/DOI-USGS/rnz/main/inst/extdata/bcsd.zarr"

z <- zarr_open(store = HttpStore$new(url))

# zarrs handles the HTTP reads + parallel decompression
pr <- z$get_item("pr")
pr
```

```{r, eval=has_zarrs && ("http_sync" %in% pizzarr_compiled_features())}
# Read a subset --- zarrs fetches only the chunks that overlap
result <- pr$get_item(list(slice(1L, 3L), slice(1L, 5L), slice(1L, 5L)))
dim(result$data)
```

### Direct zarrs_get_subset from HTTP

`zarrs_get_subset()` also works with HTTP URLs. The store handle is cached
on the Rust side, so repeated reads to the same URL reuse the connection:

```{r, eval=has_zarrs && ("http_sync" %in% pizzarr_compiled_features())}
meta <- zarrs_open_array_metadata(url, "pr")
str(meta[c("shape", "dtype", "zarr_format")])
```

```{r, eval=has_zarrs && ("http_sync" %in% pizzarr_compiled_features())}
# Read a single element (first along each dimension)
ranges <- lapply(seq_along(meta$shape), function(i) c(0L, 1L))
result <- zarrs_get_subset(url, "pr", ranges, NULL)
result$data
```

```{r, eval=has_zarrs && ("http_sync" %in% pizzarr_compiled_features())}
zarrs_close_store(url)
```

### Feature detection

Check whether HTTP support is compiled in with `pizzarr_compiled_features()`.
When `"http_sync"` is present, zarrs can open `http://` and `https://` URLs.
When it is absent, HTTP reads fall through to the R-native `crul`-based path:

```{r, eval=has_zarrs}
"http_sync" %in% pizzarr_compiled_features()
```

## S3 reads via zarrs

When the `s3` feature is compiled in, the zarrs backend can read from Amazon S3
buckets using the [object_store](https://docs.rs/object_store/) crate with an
async-to-sync adapter. Public buckets work without credentials (unsigned
requests). Authenticated access uses standard AWS environment variables
(`AWS_ACCESS_KEY_ID`, `AWS_SECRET_ACCESS_KEY`, `AWS_REGION`).

S3 stores are currently read-only via zarrs --- write operations fall through to
the R-native path.

```{r, eval=has_zarrs && ("s3" %in% pizzarr_compiled_features())}
# OME-Zarr bonsai dataset on AWS Open Data (V2, zstd, uint8)
s3_url <- "s3://ome-zarr-scivis/v0.4/64x0/bonsai.ome.zarr"

# Read array metadata
meta <- zarrs_open_array_metadata(s3_url, "scale0/bonsai")
str(meta[c("shape", "dtype", "zarr_format")])
```

```{r, eval=has_zarrs && ("s3" %in% pizzarr_compiled_features())}
# Read a small subset (first 4x4x4 corner)
result <- zarrs_get_subset(s3_url, "scale0/bonsai",
                           list(c(0L, 4L), c(0L, 4L), c(0L, 4L)), NULL)
str(result)
```

```{r, eval=has_zarrs && ("s3" %in% pizzarr_compiled_features())}
zarrs_close_store(s3_url)
```

### GCS and other cloud stores

GCS data hosted on Google Cloud Storage is publicly accessible via HTTPS
endpoints. The zarrs HTTP backend reads these directly:

```{r, eval=has_zarrs && ("blosc" %in% pizzarr_compiled_features())}
# Pangeo ECCO ocean basins (V2, blosc/lz4, float32)
gcs_url <- "https://storage.googleapis.com/pangeo-data/ECCO_basins.zarr"

meta <- zarrs_open_array_metadata(gcs_url, "basin_mask")
cat("Shape:", paste(meta$shape, collapse = " x "), "\n")
cat("Dtype:", meta$dtype, "\n")
```

```{r, eval=has_zarrs && ("blosc" %in% pizzarr_compiled_features())}
# Read a single basin mask slice
result <- zarrs_get_subset(gcs_url, "basin_mask",
                           list(c(0L, 1L), c(0L, 90L), c(0L, 90L)), NULL)
cat("Slice dimensions:", paste(result$shape, collapse = " x "), "\n")
```

```{r, eval=has_zarrs && ("blosc" %in% pizzarr_compiled_features())}
zarrs_close_store(gcs_url)
```

Authenticated GCS access via `gs://` URLs requires the `gcs` compiled feature
and GCP credentials (environment variables or application default credentials).
The `S3Store` and `GcsStore` R6 classes provide URL wrappers for high-level
use with `zarr_open()`:

```r
# S3 (requires s3 feature)
z <- zarr_open(store = S3Store$new("s3://bucket/path/to/store.zarr"))

# GCS (requires gcs feature + credentials)
z <- zarr_open(store = GcsStore$new("gs://bucket/path/to/store.zarr"))
```

## C/F order handling

zarrs stores data in C-order (row-major), while R uses F-order (column-major).
The Rust backend handles this conversion transparently:

- **Reads:** `zarrs_get_subset()` returns data in F-order, ready for
  `array(data, dim = shape)` with no `aperm()` needed.
- **Writes:** `zarrs_set_subset()` accepts F-order data and converts to C-order
  internally before writing to the store.

The transpose uses cache-blocked tiling for 2D arrays and output-order
iteration with incremental index tracking for higher dimensions, matching or
exceeding the performance of R's C-level `aperm()`.

```{r, include=FALSE}
if (clean) unlink(sample_dir, recursive = TRUE)
```
